### 2013-09-22 arfi_background.tex
final draft version

\subsection{Acoustic Radiation Force Impulse (ARFI) Imaging}
ARFI imaging is an ultrasound technique that evaluates the mechanical
properties of tissues. It generates short-duration acoustic radiation forces
that result in tissue displacement. The response can then be measured for
tissue characterization, with displacement being inversely proportional to
stiffness.~\cite{Nightingale2002} This characterization is of particular
importance in prostate evaluation due to the gland’s complex and heterogeneous
composition.  Previous studies have shown that the central zone and PCA can be
up to 3 times stiffer than PZ tissue.~\cite{Zhai2010a,Zhai2012,Zhai2010b} This
finding can be explained by the PZ’s higher water content and is analogous to
the higher signal intensity of the PZ on T2WI.

The potential of ARFI imaging to distinguish normal prostatic anatomy from
pathologic tissue is promising in the future of PCa diagnosis and image guided
therapies. B-mode US, which is currently used for real-time guidance in
TRUS-guided biopsies, does not have the ability to clearly visualize PCa or to
differentiate it from other normal structures or disease processes such as
benign prostatic hyperplasia (BPH) or prostatitis.~\cite{Zhai2010a} This has
resulted in a systematic yet random approach to prostate core biopsies, with
6-12 cores routinely sampled from different anatomical regions.~\cite{Loch2004}
The ability to clearly distinguish PCa from other tissues could lead to
targeted biopsies with increased sensitivity and resultant decreased morbidity.
It could also aid in the development of targeted therapies for those with low
burden of disease.

>>>>

\subsection{Acoustic Radiation Force Impulse (ARFI) Imaging}
Acoustic radiation force-based elasticity imaging has recently been developed
for investigation of mechanical properties of human soft tissues, including
liver, arteries, heart, prostate, and
breast~\cite{Bouchard2009,Dumont2009,Trahey2004,Dahl2009,proc_congdon04,Palmeri2011,Zhai2010a,hsu07,Palmeri2011a}.
and it provides a unique tool to use to characterize the cervix during
pregnancy.  The acoustic radiation force $\vec{F}$ generated in soft tissues by
focused ultrasound can be described by~\cite{nyborg65,torr84}:
\begin{equation}
\vec{F} = \frac{2 \alpha \vec{I}}{c},
\label{eqn:radforce}
\end{equation}
where $\alpha$ is the acoustic attenuation coefficient of the tissue, $c$ is
the tissue's sound speed, and $\vec{I}$ is the acoustic intensity at a given
point in space.  The acoustic radiation force is generated by a transfer of
momentum from the propagating acoustic wave to the propagation medium through
attenuation mechanisms, such as absorption and scattering of the ultrasonic
wave.  In these studies, the acoustic attenuation and sound speed of the
prostate tissue were assumed to be constant and not vary as a function of the
zonal anatomy, meaning that all displacement differences seen in the prostate
would be assumed to be stiffness-related after normalizing for depth-dependent
focal gain.

### 2013-09-22 mr_background.tex
final draft version

\subsection{Clinical Background and Significance}
Prostate cancer (PCa) is the most common non-cutaneous malignancy among men in
the United States. Approximately 1 in every 6 men will develop PCa during their
lifetime, with the median age of diagnosis at 67. 1It is also the second
leading cause of cancer related death, with 1 in 36 men dying from the disease.
The National Cancer Institute estimates that 238,590 men will be diagnosed with
PCa in 2013 and 29,720 will die from the disease.~\cite{Howlader2011}

PCa diagnosis usually begins by screening with prostate specific antigen (PSA)
and digital rectal examination (DRE).  Definitive diagnosis is made by random
transrectal ultrasonography-guided (TRUS) biopsies, which are then used to
provide the clinician with the proper Gleason score. The combination of these
factors—as well as staging—determines the appropriate therapy and prognosis.

PCa screening has led to earlier diagnosis of smaller tumors and more localized
disease.  However, it is well known that the sensitivity and specificity of PSA
and DRE are not optimal. In addition, DRE has a low predictive value at lower
PSA ranges, and PSA yields many false
positives.~\cite{Gosselaar2007,Gupta2013,Hricak2007} As such, a theoretical
risk of over-diagnosis and treatment of low-grade—and possibly clinically
insignificant—disease exists. Moreover, due to the random nature of biopsies,
cancer located outside the routine sampling site can be missed and the extent
of the cancer might be underestimated.~\cite{Gupta2013,Cornud1012} For example,
in a study by Mufarrij et al. 45.9--47.2\% of patients who were candidates for
active surveillance but underwent radical prostatectomy had a higher Gleason
score on final histopathology than after TRUS biopsy.~\cite{Mufarrij2010} These
inaccuracies may lead to inappropriate diagnosis, imprecise risk assessment and
potentially avoidable morbidity.

\subsection*{The Use of Magnetic Resonance Imaging in Prostate Diagnostics}
Magnetic resonance (MR) imaging has been available for use in the workup of
patients with PCa since the early 1980s but the early work on its diagnostic
accuracy is heterogeneous.7 Earlier MR techniques relied mostly on morphology
via T1 and T2- weighted imaging (T2WI). The more recent ability to include not
only anatomic but also biologic and functional dynamic parameters into MR
analysis—via diffusion-weighted imaging (DWI), dynamic contrast-enhanced (DCE)
imaging or MR spectroscopic imaging (MRSI)—is promising in the future diagnosis
and management of PCA.

Currently prostate MR focuses on a multiparametric approach, where 2 or more
imaging sequences—including anatomic and functional data—are used together to
try to arrive to a diagnosis.~\cite{Barentsz2012} As MR technology continues to evolve and
improve, its role in PCA diagnosis, staging, treatment planning and follow-up
has gained much attention.

\subsection{T2-Weighted Imaging and Prostate Anatomy}
T2WI sequences are crucial components of prostate MR imaging.  T2WI is
particularly useful in prostate analysis due to its excellent soft tissue
contrast resolution, which can be maximized by using thin sections of 3--4 mm
and a small field of view of approximately 14 cm.~\cite{Gupta2013,Bonekamp2011}
T2 sequences are the most helpful for tumor localization as they can clearly
show overall prostate morphology, internal structures and prostatic
margins.~\cite{Gupta2013}

@@

Approximately 70\% of the prostatic tissue is found in the peripheral zone,
which is high in water content and thus of higher signal intensity in
T2WI.~\cite{Jung2012} Seventy five percent of prostatic tumors are found in the
PZ and normally show hypointense T2 signal when compared to the higher
intensity PZ.~\cite{Hricak2007,Hegde2013} However, tumors can sometimes seem of
similar intensity as the surrounding tissue and false positives can occur
secondary to post biopsy changes/hemorrhage, hyperplasia or prostatitis, making
diagnosis more challenging.~\cite{Hegde2013}

\subsection{Functional MR Sequences}
Even though T2WI is the mainstay of prostate MR, its overall performance in
prostate cancer diagnosis is not optimal. The incorporation of two or more
functional sequences in multiparametric MR imaging (mpMRI) has been shown to
significantly improve the performance of MRI in cancer
diagnosis.~\cite{Turkbey2012} In fact, the European Society of Urogenital
Radiology’s (ESUR) prostate MR guidelines recommend at least 2 functional
imaging techniques in addition to T2WI in order to better characterize prostate
tumors.~\cite{Barentsz2012} In a study by Turkbey et al., researchers found
that mpMRI had a positive predictive value of 98\% in prostate cancer
detection.~\cite{Turkbey2012} Functional sequences include diffusion- weighted
imaging (DWI), dynamic contrast enhanced imaging (DCE-MRI) and MR spectroscopic
imaging (MRSI).

DWI is based on the free movement of water particles in tissue and measures the
degree of motion restriction.~\cite{Koh2007} Normal prostatic tissue is very
glandular with plenty of water molecule movement. On the other hand, tumors
have high cellular density and restricted water movement, which leads to
decreased diffusion. Due to this restricted diffusion, tumors seem to be of
higher intensity on DWI.~\cite{Koh2007}

DCE-MRI provides a measure of tumor vascularity and takes advantage of
tumor-induced angiogenesis in PCa.~\cite{Noworolski2005} Increased vascularity
in PCa results in earlier and higher peak contrast enhancement and a faster
washout when compared to normal prostatic tissue. These known characteristics
of PCa are helpful when trying to localize lesions.

Like DWI, MRSI can be helpful in lesion characterization as it provides
information on the presence of certain metabolites in
tissue.~\cite{Barentsz2012} Choline and citrate are especially useful in the
setting of PCa. Choline is critical in cell membrane synthesis and is thus
usually elevated in cells with cancerous behavior.~\cite{Gupta2013} Healthy
prostate epithelium synthesizes and secretes large quantities of citrate but
levels are decreased in PCa.~\cite{Costello1999,Kurhanewicz2002} Therefore, an
increase in the choline-to-citrate ratio on MRSI can be used as an indicator of
malignancy.~\cite{Bonekamp2011}

>>>>

\subsection{MR Imaging in Prostate Diagnostics}
MR imaging has been available for use in the workup of patients with PCa since
the early 1980s, but the early work on its diagnostic accuracy is heterogeneous
due to heavy reliance on morphology via T1 and T2-weighted imaging (T2WI). More
recent abilities to include not only anatomic, but also biologic and functional
dynamic parameters, into MR analysis via diffusion-weighted imaging (DWI),
dynamic contrast-enhanced (DCE) imaging or MR spectroscopic imaging (MRSI) is
promising in the future diagnosis and management of PCa.

Currently prostate MR focuses on a multiparametric approach, where two-or-more
imaging sequences, including anatomic and functional data, are used together to
make a diagnosis.~\cite{Barentsz2012} As MR technology continues to evolve and
improve, its role in PCa diagnosis, staging, treatment planning and follow-up
has gained much attention.

\subsubsection{T2-Weighted Imaging and Prostate Anatomy}
T2WI sequences are crucial components of prostate MR imaging.  T2WI is
particularly useful in prostate analysis due to its excellent soft tissue
contrast resolution, which can be maximized by using thin sections of 3--4 mm
and a small field of view of approximately 14 cm.~\cite{Gupta2013,Bonekamp2011}
T2 sequences are the most helpful for tumor localization, as they can clearly
show overall prostate morphology, internal structures and prostatic
margins.~\cite{Gupta2013}

@@

Approximately 70\% of the prostatic tissue is found in the PZ, which is high in
water content, and thus has higher signal intensity in T2WI.~\cite{Jung2012}
Seventy-five percent of prostatic tumors are found in the PZ and normally show
hypointense T2 signal when compared to the higher intensity
PZ.~\cite{Hricak2007,Hegde2013}; however, tumors can sometimes have similar
intensity as the surrounding tissue and false positives can occur secondary to
post-biopsy changes/hemorrhage, hyperplasia or prostatitis, making diagnosis
more challenging.~\cite{Hegde2013}

\subsubsection{Functional MR Sequences}
Even though T2WI is the mainstay of prostate MR, its overall performance in
prostate cancer diagnosis is not optimal. The incorporation of two or more
functional sequences in multiparametric MR imaging (mpMRI) has been shown to
significantly improve the performance of MRI in cancer
diagnosis.~\cite{Turkbey2012}  The European Society of Urogenital Radiology’s
(ESUR) prostate MR guidelines recommend at least 2 functional imaging
techniques, in addition to T2WI, to better characterize prostate
tumors.~\cite{Barentsz2012} In a study by Turkbey \etal, researchers found that
mpMRI had a PCa detection positive predictive value of 98\%.~\cite{Turkbey2012}
Functional sequences include diffusion-weighted imaging (DWI), dynamic contrast
enhanced imaging (DCE-MRI) and MR spectroscopic imaging (MRSI).  While mpMRI is
critical in improving PCa detection, T2WI is mainstay of delineating prostate
anatomy, and for the purposes of comparing prostate anatomy between ARFI and MR
images, T2WI were used exclusively in this study.

%DWI is based on the free movement of water particles in tissue and measures the
%degree of motion restriction.~\cite{Koh2007} Normal prostatic tissue is very
%glandular with plenty of water molecule movement. On the other hand, tumors
%have high cellular density and restricted water movement, which leads to
%decreased diffusion. Due to this restricted diffusion, tumors seem to be of
%higher intensity on DWI.~\cite{Koh2007}
%
%DCE-MRI provides a measure of tumor vascularity and takes advantage of
%tumor-induced angiogenesis in PCa.~\cite{Noworolski2005} Increased vascularity
%in PCa results in earlier and higher peak contrast enhancement and a faster
%washout when compared to normal prostatic tissue. These known characteristics
%of PCa are helpful when trying to localize lesions. 
%
%Like DWI, MRSI can be helpful in lesion characterization as it provides
%information on the presence of certain metabolites in
%tissue.~\cite{Barentsz2012} Choline and citrate are especially useful in the
%setting of PCa. Choline is critical in cell membrane synthesis and is thus
%usually elevated in cells with cancerous behavior.~\cite{Gupta2013} Healthy
%prostate epithelium synthesizes and secretes large quantities of citrate but
%levels are decreased in PCa.~\cite{Costello1999,Kurhanewicz2002} Therefore, an
%increase in the choline-to-citrate ratio on MRSI can be used as an indicator of
%malignancy.~\cite{Bonekamp2011}
